Dirace: Practical Proof Automation of Dirac Notation Equations

### 2025-01-17 labelled.tex
2025/1/17

\newcommand{\fst}{\mathsf{fst}\ }
\newcommand{\snd}{\mathsf{snd}\ }
\newcommand{\var}{\mathsf{var}\ }
\newcommand{\Var}{\mathsf{Var}\ }
\newcommand{\reg}{\mathsf{Reg}\ }
\newcommand{\DType}{\mathcal{D}}
\newcommand{\cR}{\mathcal{R}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\tD}{\tilde{D}}
\newcommand{\te}{\tilde{e}}
\newcommand{\tT}{\tilde{T}}
\newcommand{\tADD}{\widetilde{ADD}}
\newcommand{\bU}{\mathbf{U}}
\newcommand{\<}{\langle}
\newcommand{\simp}{\mathsf{Simp}}
\newcommand{\List}{\mathsf{list}}
\renewcommand{\>}{\rangle}

@@

\lz{I suggest not adding first and second. Of course, we can, decide this later after discussion.}

\lz{We further only consider constant registers.}

Assume the name set of registers $\cR$.
\begin{definition}[quantum registers]
  \begin{align*}
    R ::= r\in\cR \mid (R, R)
  \end{align*}
\end{definition}

\begin{definition}[register variable set]
The variable set of a register is defined inductively by:
\begin{itemize}
    \item $R\equiv r$: $\var R = \{r\}$;
    \item $R\equiv (R_1,R_2)$: $\var R = \var R_1 \cup\var R_2$.
\end{itemize}  
We define the following relations for quantum registers:
\begin{itemize}
  \item $R$ \textbf{belongs to} $Q$, written as $R\ \text{in}\ Q$, if $\var R\subseteq \var Q$;
  \item $R$ \textbf{is disjoint with} $Q$, written as $R \| Q$, if $\var R \cap \var Q = \emptyset$.
\end{itemize}
\end{definition}

\lz{Since the construction of the quantum register is known, the variable set can be directly computed. Similarly, we use the (computable) set as dependent parameter of the type of Labelled Dirac Term.}

@@

\begin{definition}[labelled core language]
  The \textbf{labelled core langauge} includes all symbols in the core language of Dirac notation, as well as symbols defined below.
  Here, $s\subseteq \cR$ is the set of register's name, it is computable and thus we do not introduce syntax for it.
  \begin{align*}
    T & ::= \DType(s,s) \mid \mathsf{Reg}(\sigma) \\
    e & ::= R \mid |i\>_r \mid {}_r\<i| \mid e_R \mid e_{R;R} \mid  
    e \otimes e \mid e \cdot e \\
  \end{align*}
  In other words, we don't allow variables for labelled core language for now. In particular, $|i\>_r$ and ${}_r\<i|$ are used as reserved syntax for eliminating registers.
\end{definition}

\subsubsection*{Typing rules}
\lz{context? environment? Since quantum registers are not bound variables, we might just add it to the environment.}

The extra syntax for the environment.
\begin{gather}
    E ::= \dots \mid E; r : \reg(\sigma)
\end{gather}

>>>>

\newcommand{\var}{\mathsf{var}}
\newcommand{\reg}{\mathsf{Reg}}
\newcommand{\DType}{\mathcal{D}}
\newcommand{\cR}{\mathcal{R}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\tD}{\tilde{D}}
\newcommand{\te}{\tilde{e}}
\newcommand{\tT}{\tilde{T}}
\newcommand{\tADD}{\widetilde{ADD}}
\newcommand{\bU}{\mathbf{U}}
\newcommand{\<}{\langle}
\newcommand{\simp}{\mathsf{Simp}}
\newcommand{\List}{\mathsf{list}}
\renewcommand{\>}{\rangle}

@@

Labelled Dirac notation uses register names to indicate the quantum system of vectors and operators. This allows us to talk aboud the states and operations locally, without referring to the whole system. For instance, assume $Q$ and $R$ are two registers, we have
\[
    M_Q \cdot \ket{\Phi}_{(Q, R)} = ((M \otimes I) \ket{\Phi})_{(Q, R)}.
\]
In the left hand side, the state of two subsystems is $\ket{\Phi}$, and we apply quantum operation $M$ on the system $Q$.

In this section, we introduce the language and typing of registers and labelled Dirac notations, and demonstrate how to transform the equivalence problem into the Dirac notations studied above.

We assume the set of register symbols is $\cR$. The syntax of quantum registers is defined as follows.
\begin{definition}[quantum registers]
  \begin{align*}
    R ::= r\in\cR \mid (R, R)
  \end{align*}
\end{definition}

Registers can be composed by pairs of $(R, R)$, and this structure corresponds to the structure of tensor product spaces in Dirac notations.

To reason about the registers, we define their variable set as the enumeration of all register symbols involved.

\begin{definition}[register variable set]
The variable set of a register is defined inductively by:
\begin{itemize}
    \item $R\equiv r$: $\var(R) = \{r\}$;
    \item $R\equiv (R_1,R_2)$: $\var(R) = \var(R_1) \cup\var(R_2)$.
\end{itemize}
\end{definition}

@@

\begin{definition}[labelled Dirac notation]
  The \textbf{labelled Dirac notation} includes all Dirac notation symbols and the generators defined below.
  Here, $s\subseteq \cR$ is a register variable set.
  \begin{align*}
    T & ::= \DType(s,s) \mid \mathsf{Reg}(\sigma) \\
    e & ::= R \mid |i\>_r \mid {}_r\<i| \mid e_R \mid e_{R;R} \mid
    e \otimes e \otimes \cdots \otimes e \mid e \cdot e \\
  \end{align*}
\end{definition}

$\DType(s_1, s_2)$ is the unified type for all labelled Dirac notations, where $s_1$ indicates the codomain systems and $s_2$ indicates the domain systems. For instance, labelled ket has type $\DType(s_1, \emptyset)$, and labelled bra has type $\DType(\emptyset, s_2)$.
$\reg(\sigma)$ are types for registers $R$, and the index $\sigma$ indicates the type of Hilbert space represented by the register.
Terms also include the labelled notation $e_R$ for bra, ket and $e_{R;R}$ for operators. We introduce new dot and tensor product symbols for labelled Dirac notations. In labelled Dirac notation, the structure of tensor product does not matter. Therefore $\otimes$ is an AC symbol.
Following the unified type $\DType(s,s)$, all kinds of multiplications are represented by the same dot product $e \cdot e$.
Finally, $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis for the normal form of labelled Dirac notations, where $r$ are registers symbols in $\mathcal{R}$.

\subsubsection*{Typing rules}

\[
    \frac{E[\Gamma] \vdash  \sigma : \Index}{E[\Gamma] \vdash \reg(\sigma) : \Type}
    \qquad
    \frac{E[\Gamma] \vdash r : \reg(\sigma_r) \text{ for all $r$ in $s_1$ and $s_2$} }{E[\Gamma] \vdash \DType(s_1, s_2) : \textsf{Type}} 
\]

\[
    \frac{E[\Gamma] \vdash R : \reg(\sigma)\qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K_R : \DType(\var R, \emptyset)}
    \qquad
    \frac{E[\Gamma] \vdash D : \DType(s_1,s_2)}{E[\Gamma] \vdash D^\dagger : \DType(s_2,s_1)}
\]

The typing rule for tensor product is
\[
    \frac{
        E[\Gamma] \vdash D_i : \DType(s_i,s_i') \qquad
        \bigcap_i s_i = \emptyset \qquad
        \bigcap_i s_i' = \emptyset
    }
    {E[\Gamma] \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
\]

The typing rule for dot product is
\[
    \qquad
    \frac{
        \begin{aligned}
            E[\Gamma] \vdash D_1 : \DType(s_1,s_1') \\
            E[\Gamma] \vdash D_2 : \DType(s_2,s_2')
        \end{aligned}
        \qquad 
        \begin{aligned}
            s_1 \cap s_2 \backslash s_1' = \emptyset \\
            s_2' \cap s_1' \backslash s_2 = \emptyset
        \end{aligned}
    }
    {E[\Gamma] \vdash D_1\cdot D_2 : \DType(s_1 \cup (s_2\backslash s_1'), s_2' \cup (s_1'\backslash s_2))}.
\]

\subsection{Normalization}
The main idea of normalizing labelled Dirac notations is \textit{label elimination}. It expands $e_R$ and $e_{R;R}$ by tensor products of labelled basis, which are formalized as rules below.

\begin{align*}
    & K_R \ \reduce\ \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})} (\<i_R|\cdot K). (|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n}) \\
    %
    & B_R \ \reduce\ \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})} (B\cdot |i_R\>). ({}_{r_1}\<i_{r_1}|\otimes\cdots\otimes{}_{r_n}\<i_{r_n}|) \\
    %
    & O_{R,R'} \ \reduce\ \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})}
    \sum_{i_{r'_1}\in\bU(\sigma_{r'_1})}\cdots \sum_{i_{r'_{n'}}\in\bU(\sigma_{r'_{n'}})} \\
    & \qquad (\<i_R|\cdot O\cdot |i_{R'}\>).(|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n} \otimes {}_{r'_1}\<i_{r'_1}|\otimes\cdots\otimes{}_{r'_{n'}}\<i_{r'_{n'}}|)
\end{align*}

\begin{align*}
    & \textrm{(R-SUM-PUSHD0)}
    && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
    %
    & \textrm{(R-SUM-PUSHD1)}
    && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
\end{align*}

\begin{align*}
    %
    & \textrm{(R-L-SORT0)}
    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
    %
    & \textrm{(R-L-SORT1)}
    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
    %
    & \textrm{(R-L-SORT2)}
    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m) \\
    %
\end{align*}
